Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-05-14 endofquarterpaper.qmd
modularity codee

If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase. We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table. None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.

@@

Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta]. We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017]. Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorothm. Results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths, and network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.

@@

One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites. The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994]. It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.

It is therefore possible that some of the sites shown in @fig-div-plot which have a more diverse range of sign types than predicted by sample size, such as Les Cottés in the Proto-Aurignacian, Solutré in the Early Aurignacian, and Bockstein-Törle and Vogelherd in the Evolved Aurignacian acted as aggregation sites, or served a similar purpose.

Variation in diversity of sign types may indicate variation in social structure. Despite the tendency to conceive of hunter-gatherers as homogenous, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006]. Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982]. It is therefore possible that greater diversity in geometric signs indicates a wider variety of formal social roles [@Conkey_1985]. This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formal social structure, and the restricted groups having a less formal social structure.

@@

  ## ggplot
  ggplot2::ggplot(data = count) +
    ggplot2::aes(x = .data$x, y = .data$y, label = .data$label) +
    ggplot2::geom_point() +
    gg_sim +
    ggplot2::scale_x_log10(name = "Sample size") +
    ggplot2::scale_y_continuous(
      name = sprintf("%s (%s)", y_lab, object@method)
    )
}

@@

@@

# Conclusion
This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].

One limitation is the potential ambiguity that arises with some of the sign types. While @Von_Petzinger_2009 argues in favor of Aurignacian geometric signs as symbolic expression, potential ambiguity still remains over whether function is consistent across all sign type occurrences, particularly in regards to the simpler sign types composed from fewer elements. We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023]. This ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains more detailed information on the specific order and patterns in which signs occurred on individual objects.

>>>>

If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase. We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table. None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.

@@

Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta]. We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017]. Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorothm. To avoid cluttering the graph, all edges below the value of 0.2 were removed. Results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths, and network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.

@@

One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites. The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].

One potential identifier of aggregation sites comes from @fig-div-plot, which uses the tabula package [@Frerebeau_2025] to generate the Shannon-Weaver diversity index of sign types per site, and plot the expected diversity index for a given sample size against the actual diversity index per site [@Kintigh_1984]. The plot identifies multiple sites that diverge from the diversity index predicted by sample size, including those with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian. It is possible that those sites with a greater than expected diversity index acted as aggregation sites or served a similar purpose. @fig-neighbor-net also shows La Ferrassie in the Proto-Aurignacian as having multiple connections to sites in both groups, perhaps suggesting that it served as an inter-group aggregation site.

@@

  ## ggplot
  ggplot2::ggplot(data = count) +
    ggplot2::aes(x = .data$x, y = .data$y, label = .data$label) +
    ggplot2::geom_point() +
    gg_sim +
    ggplot2::scale_x_log10(name = "Sample size") +
    ggplot2::scale_y_continuous(name = "Diversity")

}

@@

@fig-div-plot shows a great deal of general divergence from the expected diversity index for the sample size. Another potential explanation for the observed variation in diversity is that it may indicate variation in social structure. Despite the tendency to conceive of hunter-gatherers as homogenous, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006]. Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982]. It is therefore possible that greater diversity in geometric signs indicates a wider variety of formal social roles [@Conkey_1985]. This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formal social structure, and the restricted groups having a less formal social structure.

@@

# Conclusion

This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].

One limitation is the potential ambiguity that arises with some of the sign types. While @Von_Petzinger_2009 argues in favor of Aurignacian geometric signs as symbolic expression, potential ambiguity still remains over whether function is consistent across all sign type occurrences, particularly in regards to the simpler sign types composed from fewer elements. We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, even if it is random copying [@Herskind2023], and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023]. This ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains more detailed information on the specific order and patterns in which signs occurred on individual objects.
